---
title: "HomeWork 3, CHASE MAUGHAN u0962361"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 2  # upto three depths of headings (specified by #, ## and ###)
    toc_float:
      collapsed: false
      smooth_scroll: false
    df_print: paged
  # 
  # number_sections: true  ## if you want number sections at each table header
  # theme: united  # many options for theme, this one is my favorite.
  # highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# HomeWork 3

## 1.

### A biomedical informatics researcher is investigating the relationship between smoking and multiple biological measurements. We want to compare the mean age for both smokers and non-smokers and we want to sample participants from each group proportionally to their representation in the population. The dataset smoking.csv contains this information, with two columns: "age" (numeric) and "smoking" (factor with 2 levels "0" and "1" [yes/no]). Answer the following questions using the concepts seen in class.

#### 1. **Point Estimate:**

#### Calculate the mean age for the entire population based on the collected sample.

```{r}
print(getwd())

```

```{r}
smoking <- read.csv("Stats and Prob/smoking.csv", header = TRUE)
age_mean <- mean(smoking$age)
print(age_mean)

```



#### 2. **Random Sampling:**

#### Randomly select a subset of 50 patients from the dataset without replacement. Calculate the mean age for this subset.


```{r}
random_sample <- smoking[sample(nrow(smoking),50, replace = FALSE),]
rs_mean <- mean(random_sample$age)
print(rs_mean)

```



#### 3. **Resampling:**

#### Perform bootstrapping on the entire dataset to estimate the sampling distribution of the mean age for the cohort Use 1000 bootstrap samples and calculate the mean age for each sample.



```{r}
n <- 1000

bootstrap_means <- numeric(n)

for (i in 1:n) {
  bootstrap_sample <- smoking[sample(nrow(smoking), replace = TRUE), ]
  bootstrap_means[i] <- mean(bootstrap_sample$age)
}


pop_mean <- mean(bootstrap_means)

print(paste("Mean of Bootstrap samples: ", pop_mean))

```




#### 4. **Confidence Intervals:**

#### Calculate a 95% confidence interval for the population mean age level using the bootstrap distribution obtained in the previous step.


```{r}
confidence_interval <- quantile(bootstrap_means, c(0.025, 0.975))

print(paste("95% Confidence Interval: [", confidence_interval[1], ", ", confidence_interval[2], "]"))

```



#### 5. **Standard Error of the Mean (SEM):**

#### Calculate the standard error of the mean (SEM) of your estimate.



```{r}
se_m <- sd(bootstrap_means)

print(paste("Standard Error of the Mean: ", se_m))


```

## 2.

### Markov Chain: physical exercise training method A is used only 5% of the time, a person using method A will stay with this method 85% of the time, and a person not using method A will switch to method A about 65% time. At the beginning of the experiment only 5% of people used method A.

#### 1. Generate a transition matrix for this Markov chain




```{r}
stay_A <- 0.85  # probability of staying with method A when already using it
switch_to_A <- 0.65  # probability of switching to method A when not using it
switch_to_B <- 0.15  # probability of switching from method A to not using it
stay_B <- 0.35  # probability of staying with not using method A

transition_matrix <- matrix(c(stay_A, switch_to_A, switch_to_B, stay_B), nrow = 2, byrow = TRUE,
                            dimnames = list(c("A", "B"), c("A", "B")))

# Display the transition matrix
print(transition_matrix)
                          
```



#### 2. Generate a transition plot (using R or by hand as an image it’s valid)

```{r}

library(igraph)

# Create a graph from the transition matrix
graph <- graph_from_adjacency_matrix(transition_matrix, mode = "directed", weighted = TRUE)

# Plot the graph
plot(graph, layout = layout.circle, vertex.label = colnames(transition_matrix),
     edge.label = round(E(graph)$weight, 2), main = "Markov Chain Transition Diagram")
                          
```

#### 3. Plot the change in the probabilities over time for both methods until the 10th time unit.
```{r}
# Set initial probabilities
initial_probabilities <- c(0.05, 0.95)  # 5% chance to start with A, 95% chance to start with B
current_probabilities <- initial_probabilities

# Set up a matrix to store the probabilities over time
probabilities_over_time <- matrix(0, nrow = 10, ncol = 2)
probabilities_over_time[1,] <- current_probabilities

# Simulate the change in probabilities over time
for (t in 2:10) {
  current_probabilities <- current_probabilities %*% transition_matrix
  probabilities_over_time[t,] <- current_probabilities
}

# Plot the results
time_points <- 1:10
plot(time_points, probabilities_over_time[,1], type='l', col='blue', ylim=c(0,1), xlab='Time', ylab='Probability', main='Probabilities over Time')
lines(time_points, probabilities_over_time[,2], type='l', col='red')
legend("topright", legend=c("A", "B"), col=c("blue", "red"), lty=1)

                          
```

## 3.

### Random Walk: Another simpler example of a random walk is a one-dimensional random walk. first we place a marker at zero (our initial state), we flip a coin, if it lands on heads, the marker is moved one unit to the right (1), if it lands on tails it is moved one unit to the left.

#### 1. Generate a function that randomly draws from our initial state and populates a vector with the different transitions.

```{r}

generate_random_walk <- function(num_steps) {
  position <- 0
  transitions <- numeric(num_steps)
  
  for (step in 1:num_steps) {
    coin_flip <- sample(c(-1, 1), size = 1)
    position <- position + coin_flip
    transitions[step] <- coin_flip
  }
  
  return(list(transitions = transitions, end_position = position))
}


random_walk_result <- generate_random_walk(10)

print(random_walk_result)

```

#### 2. Generate a plot that shows 500 independent one-dimensional walks, differentiating walks that end above 0 or below 0.

```{r}

num_walks <- 500
num_steps <- 50

# Generate random walks

walks <- lapply(1:num_walks, function(i) generate_random_walk(num_steps))

# Plot the results
plot(1, type = "n", xlim = c(1, num_steps), ylim = c(min(sapply(walks, function(w) min(cumsum(w$transitions)))), 
                                                      max(sapply(walks, function(w) max(cumsum(w$transitions))))), 
     xlab = "Steps", ylab = "Position", main = "500 Independent One-Dimensional Walks")

for (i in 1:num_walks) {
  end_position <- walks[[i]]$end_position
  color <- ifelse(end_position > 0, "blue", "green")
  lines(1:num_steps, cumsum(walks[[i]]$transitions), col = color, lwd = 0.8)
}


```

#### 3. What is the frequency of walks that ended in a positive cumulative count, in zero, or negative?

```{r}

positive <- sum(sapply(walks, function(w) w$end_position > 0))
zero <- sum(sapply(walks, function(w) w$end_position == 0))
negative <- sum(sapply(walks, function(w) w$end_position < 0))


cat("Frequency of walks ending in a positive cumulative count:", positive, "\n")
cat("Frequency of walks ending in zero:", zero, "\n")
cat("Frequency of walks ending in a negative cumulative count:", negative, "\n")


```
